On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification

### 2021-03-11 paper/cav_determinisim_main.tex
GC up to Sec 7

\subsubsection{Simulation Variance \& Deviation}
If the simulator is non-deterministic then how can the simulation variance be measured? This can be achieved by monitoring the values of any of the recorded output variables that should be consistent from run to run. 
% In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
For example, actor position variance is a distance-based metric that can be derived from simulation traces. The actor position over time, i.e.\ the actor path, is often used in assertion checking, e.g.\ to determine whether vehicles keep within lanes or whether minimum distances to other vehicles and road users are being maintained. 
% 
Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor position variance in the simulation with respect to time, assuming fixed actions. {\color{red} Unclear: Why wrt.\ time, that would then be the actor path, or not?}
%
%\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive measure to comprehend when interpreting test results.
%
Case study results are presented using deviation (SI unit $m$), the square root of variance, rather than variance, as this is a more intuitive measure to comprehend when interpreting test results.
%
% more intuitive to comprehend when considering verification tests, e.g. vehicle close-passing. Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.\\
\\

@@

% In pursuit of an answer to this question we
In this paper we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
%
In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
% KIE: This is covered in the next sentence:
% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
%
\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified. }
%
% By analysing actor position variance 
We find that the CARLA simulator is non-deterministic under certain conditions. {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
%
CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. 
%
\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}

@@

\subsection{Non-Uniform Memory Access (NUMA)}
For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor. Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}. 
%
Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically if tasks are processed out of sequence using equal priority scheduling, or, perhaps, simply with an increased data transfer cost, i.e.\ slower. 
%
By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.

@@

\subsection{Summary}
We have investigated the potential sources of non-determinism affecting game engines 
and explored the impact they may have on simulation variance. 
%
% \todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable. }
%
Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
\todo[inline]{So, do commercial HPC facilities have ECC memory?}
\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
%
\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.}
To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
%
Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer. Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC. {\color{red} Not sure this is correct, see the NUMA section above. Why ``only''? If one computation out of many takes longer, then their exec order can be affected, or not?}
%
Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups. 
%
However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop. Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism. 
%

@@

To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
%
Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
%
% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
%
The tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small as to fail with minor computational perturbations. {\color{red} {\sc What would fail?}}
%
Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.

@@

Instead of using ..., {\color{red} \sc please complete} herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
%
\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
%
Repeated traces are compared at the same time index $t$ for each actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
%
Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation. }

@@

% ***************************************************
%  Internal Settings
% ***************************************************
\subsection{Internal Settings}

@@

\subsection{Initial Results and Discussion}\label{InitialResultsSection}

{\color{red} This is for unrestricted, so this does not come as a surprise later.}

@@

After discovering that internal engine settings could not achieve a simulation variance that met the required 
% \textit{deterministic threshold}, 
\textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED? Should there be an ``Interim Results'' section?}
settings external to the game engine were explored. 
%
Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment. 
%
For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}. This hypothesis was supported by our initial work performed on the unreal engine~\cite{TSLUnrealEngineTesting} and was explored more fully in this work using the CARLA platform. {\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented. Either take out or include, at least wrt.\ the overall observations, such as the support for this hypothesis.}

To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation. 
%
% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0 to 95\% (see Section~\ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively. Resource utilisation figures reported here should be considered approximate values.
% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively). 
%
{\color{red} Somewhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}

@@

% ***************************************************
%  System Configuration and Screening
% ***************************************************
\subsection{Experimental System Configuration and Pre-Screening}\label{s:screening}
The experiments were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
%
The operating systems was Linux Ubuntu 18.04.4 LTS. 
%
Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s. 
%
Each test was repeated 1000 times. {\color{red} Justify why 1000 - see my later comment in the Method section.}
%
A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
%
To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system. These were:

\begin{itemize}[leftmargin=*]
    \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, \cite{shi2009testing}.
    \item Non-uniform memory access (NUMA): \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores, but only a minor (2\%) improvement in simulation variance was observed. Therefore this was not used for subsequent testing. {\color{red} But if we aim to min the sources of non-det, then that should be considered/used?}
\end{itemize}

@@

@@

% ***************************************************
%  Results and Discussion
% ***************************************************
\subsection{Final Results and Discussion}\label{FinalResultsSection}

{\color{red} Consider spitting to introduce the first part of the table, which motivates the 2nd lot of experiments under load restrictions. I'd find this more logical.}

@@

% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton. Note the dotted red line in the plot shows the millimeter (mm) level. Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable. This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
%
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v6.pdf}
    \caption{Stages of the method proposed to determine the variance of a simulation. {\color{red} \sc please update to match text}}
    \label{method_diagram}
\end{figure*}

@@

\subsection{Process Scheduling Priority}
An investigation into the impact of process scheduling on the simulation variance was undertaken {\color{red} say why}. The experiment was repeated ($n=1000$) using Test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
%
To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.

The results are presented in Fig.~\ref{NICEExperimentStressSummary} {\color{red} What is on the y axis?} where the box denotes the inter-quartile range {\color{red} of???}, non-outlier limits by the whiskers and a horizontal bar for the median. The figure shows that decreasing the priority {\color{red} of the simulator process} (right hand side of the  plot) has little effect on simulation variance when compared to a default NICE value of 0 (central bars in the plot). Increasing priority (left hand side of the plot) significantly reduced the variance for the 75\% resource utilisation experiment, but this does not account for all the difference in the observed results. This can be seen in the maximum priority setting where the bars in the plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling. 
%
The remaining difference in the variance between the two resource utilisation levels may be due to the lack of absolute control that \texttt{NICE} has over process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}

\subsection{Investigation Summary} \label{s:empirical_summary} {\color{red} Factor into overall conclusion of the case study}

% \todo[inline]{GC-need to reference the 'restricted' set of results here too}

%\todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be?? The results indicate that actor type, resource utilisation and collision events all affect the determinism in the CARLA simulator. Collision events may utilise more computational resources than simulations without such events, i.e. the physics loop takes longer to process as there are complex physical interactions to resolve. This may coincide with the second observation that higher resource levels also lead to non-deterministic results. We hypothesise that both these observations are due to improper management of thread scheduling within the engine. When resources are occupied there may be a delay in processing critical calculations for the simulator, which may arrive out of an expected sequence or the engine compensates for this delay by reducing the number of physics loops that can be performed within a render cycle. (is this 2nd one true, or does carla ignore render requests?)}

This empirical investigation has highlighted the shortcomings of using a games engine for simulation based verification and advice for best working practice. However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly. 
%
Therefore we have derived a general methodology that practitioners can follow to find the 
% \textit{deterministic domains of operation} 
\textit{operational domains of permissible variance} 
for %any hardware configuration.
a game-engine-based simulation environment. This methodology is presented in the next section.

@@

% ***************************************************
%  METHOD
% ***************************************************
\section{Method to determine the Variance of a Simulation {\color{red} Say for what???}} \label{s:methodology}
In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}. This method can be used to determine the operational domains of permissible variance of a simulation environment.  
%
{\color{red} MAKES NO SENSE: {\sc In this section a method for determining the variance of simulated actor path trajectories (KIE: a trajectory is a path), termed simulation variance, and resolving the operational domains of permissible variance of the simulation presented here as a work flow, see Fig.~\ref{method_diagram}.}}
%
In addition, recommendations and best practice guidelines for minimising simulation variance are suggested. {\color{red} \sc Highlight these in the text.}
% A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
% This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}. %These need to be considered at different stages in the process. 
%

@@

\subsubsection{Sample Size}
% Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo[inline]{Explain what this means.} 
% Histogram to show how sampling with position binning varies?
% \footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}
%
Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm. While executing 100 repeats may seem sufficient, this sample size may fail to observe events that occur with low probability, giving false confidence in the results. {\color{red} \sc Again, separate the study from the method. This observation should really have been made in the section on ``Experimental System Config and Pre-Screening, to justify the figure of 1000 repeats.} It is therefore recommended that the sample size is determined empirically dependent on the number of results that exceed the tolerance. {\color{red} \sc How would one know?}
% To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.

@@

\subsection{Analysis}
For each experiment, the maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control to reach the domain of permissible variance within the simulation environment. 
%Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.
% KIE: Why now point to HW? And, which "boundaries"?

>>>>

\subsubsection{Simulation Variance \& Deviation}
If the simulator is non-deterministic then how can the simulation variance be measured? This can be achieved by monitoring the values of any of the recorded output variables that should be consistent from run to run. 
% In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
For example, actor position variance is a distance-based metric that can be derived from simulation traces. The actor position over time, i.e.\ the actor path, is often used in assertion checking, e.g.\ to determine whether vehicles keep within lanes or whether minimum distances to other vehicles and road users are being maintained. 
% 
Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor path variance in the simulation, assuming fixed actions. 
% {\color{red} Unclear: Why wrt.\ time, that would then be the actor path, or not?}
%
%\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive measure to comprehend when interpreting test results.
%
Case study results are presented using deviation (SI unit $m$), the square root of variance, rather than variance, as this is a more intuitive measure to comprehend when interpreting test results.
%
% more intuitive to comprehend when considering verification tests, e.g. vehicle close-passing. Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.\\
\\

@@

\subsection{Contribution}
% In pursuit of an answer to this question we
In this paper we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
%
In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
% KIE: This is covered in the next sentence:
% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
%
\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified. }
%
% By analysing actor position variance 
We find that the CARLA simulator is non-deterministic. Actor path variance was found to be non-zero under certain conditions where the deviation from the mean was observed up to $59$cm. We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment. 
% {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
%
CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. 
%
\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}

@@

\subsection{Non-Uniform Memory Access (NUMA)} \label{s:sources_numa}
For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor. Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}. 
%
Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically if tasks are processed out of sequence using equal priority scheduling, or, perhaps, simply with an increased data transfer cost, i.e.\ slower. 
%
By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.

@@

\subsection{Summary}
We have investigated the potential sources of non-determinism affecting game engines 
and explored the impact they may have on simulation variance. 
%
% \todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable. }
%
Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
\todo[inline]{So, do commercial HPC facilities have ECC memory?}
\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
%
\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.}
To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
%
Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer. Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC. If this change in computational cycle gives opportunity for the execution order to be changed then this situation may lead to non-determinism.

% {\color{red} Not sure this is correct, see the NUMA section above. Why ``only''? If one computation out of many takes longer, then their exec order can be affected, or not?}
%
Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups. 
%
However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop. Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism. 
%

@@

To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
%
Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
%
% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
%
The tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small such that assertion checking would fail with minor computational perturbations. 
% {\color{red} {\sc What would fail?}}
%
Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.

@@

Instead of using variance, herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor path relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
%
\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
%
Repeated traces are compared at the same time index $t$ for each actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
%
Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation. }

@@

% ***************************************************
%  Internal Settings
% ***************************************************
\subsection{Simulator Settings}

@@

% \subsection{Initial Results and Discussion}\label{InitialResultsSection}

% {\color{red} This is for unrestricted, so this does not come as a surprise later.}

@@

% After discovering that internal engine settings could not achieve a simulation variance that met the required 
% \textit{deterministic threshold}, 
% \textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED? Should there be an ``Interim Results'' section?}
After examining controls the internal game engine, external settings were explored. 
%
Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment. 
%
For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.

Some initial exploratory work was undertaken~\cite{TSLUnrealEngineTesting} that suggested computational resource utilisation was positively correlated with simulation variance, leading to high simulation variance when a system is under high load. This initial observation was explored more fully in this work using the CARLA platform.

% {\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented. Either take out or include, at least wrt.\ the overall observations, such as the support for this hypothesis.}

To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation. 
%
% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0 to 95\% (see Section~\ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively. Resource utilisation figures reported here should be considered approximate values. %
Resource utilisation was capped to 75\% in some parts of the results and referred to as \textit{restricted}. This was done to limit the load on the test system and hence limiting simulation variance. The term \textit{unrestricted} places no such limit on the system load.

% {\color{red} To understand the impact of collisions and high resource utilisation, the restricted column shows a subset of the results where post-collision data and experiments above 75 resource utilisation have been removed}
%
% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively). 
%
% {\color{red} Somewhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}

@@

% ***************************************************
%  System Configuration and Screening
% ***************************************************
\subsection{Experimental System Configuration and Pre-Screening}\label{s:screening}
The experiments were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
%
The operating systems was Linux Ubuntu 18.04.4 LTS. 
%
Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.

% Each test was repeated 1000 times. 
Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm. While executing 100 repeats may seem sufficient, this sample size may fail to observe events that occur with low probability, giving false confidence in the results. 
Therefore each test was repeated 1000 times to provide a sufficient sample size.
% {\color{red} Justify why 1000 - see my later comment in the Method section.}
%
A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
%
To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system. These were:

\begin{itemize}[leftmargin=*]
    \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, \cite{shi2009testing}.
\end{itemize}

@@

% ***************************************************
%  Results and Discussion
% ***************************************************
\section{Results and Discussion}\label{FinalResultsSection}

@@

% {\color{red} Consider spitting to introduce the first part of the table, which motivates the 2nd lot of experiments under load restrictions. I'd find this more logical.}

@@

% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton. Note the dotted red line in the plot shows the millimeter (mm) level. Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable. This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
%
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v7.pdf}
    \caption{Stages of the method proposed to determine the variance of a simulation.}
    \label{method_diagram}
\end{figure*}

@@

\subsection{Process Scheduling Priority}

An investigation into the impact of process scheduling on the simulation variance was undertaken following the observations of increased simulation variance with increased resource utilisation. The experiment was repeated ($n=1000$) using Test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
%
To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.

The results are presented in Fig.~\ref{NICEExperimentStressSummary} where the box denotes the inter-quartile range of deviation, non-outlier limits by the whiskers and a horizontal bar for the median. The figure shows that decreasing the priority of the simulator process (right hand side of the  plot) has little effect on simulation variance when compared to a default NICE value of 0 (central bars in the plot). Increasing priority (left hand side of the plot) significantly reduced the variance for the 75\% resource utilisation experiment, but this does not account for all the difference in the observed results. This can be seen in the maximum priority setting where the bars in the plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling. 
%
The remaining difference in the variance between the two resource utilisation levels may be due to the lack of absolute control that \texttt{NICE} has over process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}

\subsection{Non-uniform memory access (NUMA)}
An additional investigation into memory access was undertaken given the potential impact that data transfer cost and execution order may have on simulation variance, see Section~\ref{s:sources_numa}. %
The program \texttt{numactl}~\cite{numactl_NUMA} allows a process to run with a specified memory placement policy, essentially allowing the process to be bound to a particular CPU core. %
\texttt{numactl} was used to fix the simulator and test script to single cores, and a (2\%) improvement in simulation variance was observed. %
This was considered minor comparative to the changes in simulation variance observed in other aspects of the case study, for example the pre- and post-collision between a vehicle and pedestrian saw a change of $10^{14}$. Therefore \texttt{numactl} was not used for subsequent testing.

However, practitioners that are aiming to minimise simulation variance should use a technique such as this to minimise data transfer cost and potentially minimise any potential effect on process execution order that may lead to increased simulation variance.

\subsection{Investigation Summary} \label{s:empirical_summary} 
% {\color{red} Factor into overall conclusion of the case study}
% \todo[inline]{GC-need to reference the 'restricted' set of results here too}

%\todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be?? The results indicate that actor type, resource utilisation and collision events all affect the determinism in the CARLA simulator. Collision events may utilise more computational resources than simulations without such events, i.e. the physics loop takes longer to process as there are complex physical interactions to resolve. This may coincide with the second observation that higher resource levels also lead to non-deterministic results. We hypothesise that both these observations are due to improper management of thread scheduling within the engine. When resources are occupied there may be a delay in processing critical calculations for the simulator, which may arrive out of an expected sequence or the engine compensates for this delay by reducing the number of physics loops that can be performed within a render cycle. (is this 2nd one true, or does carla ignore render requests?)}
These empirical investigations have highlighted the shortcomings of using a games engine for simulation based verification and suggests advice for best working practice. 
%
It was observed that resource utilisation positively correlates with simulation variance and that specific simulation events, such as vehicle collisions, can also lead to a breach in permissible tolerance. 
%
The investigation found that the effect of higher simulation variance as a result of increased resource utilisation can be reduced, but not omitted entirely, by controlling the scheduling policy. 
%
The investigation into specifying memory placement on the simulation process did improve simulation variance but only by a minor amount of 2\%.

However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly. 
%
Therefore we have derived a general methodology that practitioners can follow to find the 
% \textit{deterministic domains of operation} 
\textit{operational domains of permissible variance} 
for %any hardware configuration.
a game-engine-based simulation environment. This methodology is presented in the next section.

@@

% ***************************************************
%  METHOD
% ***************************************************
\section{Method to determine the Variance of a Simulation} \label{s:methodology}
% In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}. This method can be used to determine the operational domains of permissible variance of a simulation environment.  
%
In this section a method for determining the simulation variance of actor paths and resolving the operational domains of permissible variance is presented here as a work flow, see Fig.~\ref{method_diagram}.
%
In addition, recommendations and best practice guidelines for minimising simulation variance are suggested, such as fixing the physics time step in \textit{Simulator Settings}. 
% {\color{red} \sc Highlight these in the text.}
% A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
% This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}. %These need to be considered at different stages in the process. 
%

@@

\subsubsection{Sample Size}
% Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo[inline]{Explain what this means.} 
% Histogram to show how sampling with position binning varies?
% \footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}
%
% {\color{red} \sc Again, separate the study from the method. This observation should really have been made in the section on ``Experimental System Config and Pre-Screening, to justify the figure of 1000 repeats.} 
It is recommended that the chosen sample size is determined empirically. By increasing the sample size and observing the simulation variance does not change
...
...
 dependent on the number of results that exceed the tolerance. {\color{red} \sc How would one know?}
% To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.

@@

\subsection{Analysis}
For each experiment, the maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control to reach the domain of permissible variance within the simulation environment.

%Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.
% KIE: Why now point to HW? And, which "boundaries"?
